---
title: "Breast Cancer Type Prediction using Proteomic Data"
author: "Ngoc Duong - nqd2000"
date: "05/15/2020"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(tidyverse)
library(data.table)
library(viridis)
library(mgcv)
library(ggplot2)
library(pdp)
library(earth)
library(patchwork)
library(nnet)
library(janitor)
library(FNN)
library(ModelMetrics)
library(caret)
library(boot)
library(microbenchmark)
library(broom)
library(corrplot)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

Import data and get rid of NA values
```{r}
#import data
outcome = read.csv("./breastcancerproteomes/clinical_data_breast_cancer.csv") %>% janitor::clean_names()
proteome = read.csv("./breastcancerproteomes/77_cancer_proteomes_CPTAC_itraq.csv") 

#clean proteome data
#transpose dataset
proteome_tp <- transpose(proteome)

#get row and colnames in order
colnames(proteome_tp) <- proteome$RefSeq_accession_number
proteome_tp$par_id <- colnames(proteome) 

#rearrange data
proteome_with_id = as_tibble(proteome_tp) %>% select(par_id, everything()) %>% .[-c(1:3),] %>% separate(par_id, c("id2","id4","tcga")) %>% select(-tcga)

proteome_wo_id = as_tibble(proteome_tp) %>% select(-par_id, everything()) %>% .[-c(1:3),-ncol(proteome_tp)]
  
#clean outcome data 
outcome_clean = outcome %>% 
  separate(complete_tcga_id, c("tcga","id2","id4"), "-") %>% #sep id based on 2-digit id and 4-digit id
  select(-tcga) %>% 
  select(id2, id4, node_coded)

bcp_merge = left_join(proteome_with_id, outcome_clean, by = c("id2","id4"))
```

Leave out proteins that have missing quantification values 

```{r}
missing.counts <- NULL
for(i in 1:ncol(proteome_wo_id)) {
missing.counts[i] <- sum(is.na(proteome_wo_id[,i]))}

miss <- names(proteome_wo_id)[which(missing.counts >= 1)] 

proteome_woid_nona = proteome_wo_id[ ,!(colnames(proteome_wo_id) %in% c(miss))]
```

Screening variables 

```{r}
#find variance for all proteins
var <- NULL
for(i in 1:ncol(proteome_woid_nona)){
var[i] <- var(proteome_woid_nona[,i])}

var_id = tibble(row = c(1:7994), var) %>% arrange(desc(var)) %>% slice(1:500)

#find names of proteins with high variances and put them in a vector
sup.var = names(proteome_woid_nona)[var_id$row]

#subset original data with the created vector
proteome_final = proteome_woid_nona[,c(sup.var)]
```


Obtain final data
```{r}
#merge proteome and clinical data 
bcp_data = cbind(proteome_final, node = bcp_merge$node_coded) %>% as_tibble() %>% drop_na(node) %>% mutate_at(vars(-node), as.numeric) 
```

Exploratory data analysis

```{r}
#grouped boxplots
#top 20 genes with highest variance
bcp_data %>% dplyr::select(c(1:20, 501)) %>% 
  pivot_longer(1:20,
               names_to = "protein",
               values_to = "value") %>% 
  group_by(node) %>% 
  ggplot(aes(x = protein, y = value, color = node)) +
  geom_boxplot() + theme_bw() + 
  labs(y = "Expression levels", x = "Protein") + facet_grid(~node) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))

#mid 30 proteins with highest variance
bcp_data %>% dplyr::select(c(91:110, 501)) %>% 
  pivot_longer(1:30,
               names_to = "protein",
               values_to = "value") %>% 
  group_by(node) %>% 
  ggplot(aes(x = protein, y = value, color = node)) +
  geom_boxplot() + theme_bw() + 
  labs(y = "Expression levels", x = "Protein") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))

#40 proteins with lowest variance
bcp_data %>% dplyr::select(c(161:200, 501)) %>% 
  pivot_longer(1:40,
               names_to = "protein",
               values_to = "value") %>% 
  group_by(node) %>% 
  ggplot(aes(x = protein, y = value, color = node)) +
  geom_boxplot() + theme_bw() + 
  labs(y = "Expression levels", x = "Protein") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))
```


## Prediction

```{r}
#create training set with a random sample of 800 observations
set.seed(2020)
rowTrain <-createDataPartition(y = bcp_data$node,
                               p = 0.75,
                               list = FALSE)
bcp_train = bcp_data[rowTrain,]
bcp_test = bcp_data[-rowTrain,]
```


### Random Forest

```{r}
ctrl_cl = trainControl(method = "repeatedcv",
                    summaryFunction = twoClassSummary,
                    classProbs = TRUE)

rf.grid = expand.grid(mtry = 15:70, 
                      splitrule = "gini",
                      min.node.size = 1:5)

set.seed(2020)
rf.fit <- train(node~., bcp_train,
                method = "ranger",
                tuneGrid = rf.grid,
                metric = "ROC",
                trControl = ctrl_cl, 
                verbose = FALSE)

ggplot(rf.fit, highlight = TRUE)
```

Random Forest prediction results

```{r}
rf.pred = predict(rf.fit, bcp_test, type = "prob")
rf.pred
```


### Boosting 

```{r}

```


### SVM

**Fit a support vector classifier (linear kernel) to the training data with Tumor Type as the response**

```{r}
ctrl <-trainControl(method = "repeatedcv", number = 5, repeats = 5)

set.seed(2020)
svml.fit <-train(node~.,data = bcp_train,
                 method = "svmLinear2",
                 allowParallel=TRUE,
                 tuneGrid =data.frame(cost =exp(seq(-10,2,len=30))),
                 trControl = ctrl)

ggplot(svml.fit, highlight = TRUE)
```

Find the training and test error rate

```{r}
pred.svml.test <-predict(svml.fit, newdata = bcp_test)
summary(pred.svml.test)
table(pred.svml.test, bcp_test$node)
confusionMatrix(data = summary(pred.svml.test), reference = bcp_test$node)
```

**Fit a support vector machine (radial kernel) to the training data with Tumor Type as the response**

```{r}
svmr.grid <-expand.grid(C =exp(seq(-5,2,len=10)),
                        sigma =exp(seq(-7,1,len=10)))

set.seed(7)
svmr.fit <-train(tumor~.,data = bcp_train,
                 method = "svmRadial",
                 allowParallel=TRUE,
                 tuneGrid = svmr.grid,
                 trControl = ctrl)

ggplot(svmr.fit, highlight = TRUE)
```

Find the training and test error rate

```{r}
pred.svmr.test <-predict(svmr.fit, newdata = bcp_test)
table(pred.svmr.test, bcp_test$tumor)
```

